Comparison of Spatial Partitioning Strategies

### 2025-02-28 paper/paper.tex
paper: Change the title, abstraction and wrote the last analysis parts.

    {\Huge \textbf{Comparison of Spatial Partitioning Strategies}}\\[1.5cm]

@@

\begin{abstract}
    Spatial partitioning is a widely used optimization strategy to efficiently
    optimize objects in space. It can be found in many applications, from
    simulating galaxies to video games.
    Checking collision between objects is an expensive operation that quickly
    becomes a bottleneck as the number of objects grow.
    This paper compares three different strategies, alongside a naive solution.
    The aim is to evaluate the use cases for the different strategies.
    Each strategy was implemented to fit the same interface to make it as fair
    of a comparison as possible.
    The results show how spatial hashing out performs every other method when
    the objects are distributed evenly while quadtrees excel with uneven object
    distribution.
\end{abstract}

@@

@@

\section{Results \& Analysis}
This section presents the results and offers an analysis of the benchmarking
done on the three strategies - quadtrees, spatial hashing and static grids. The
metrics measured was average runtime: the sum of all operations needed to run
the program. Insertions: how long it took to insert all the objects into the
data structure. Collision testing: total time spent checking objects against
each other, which includes querying. Querying: average time spent retrieving
objects from the data structure. Clearing: how long it takes to reset the data
structure making it ready for rebuilding. Each metric was measured with a range
of object count, labeled as 'box count' in the graphs, from ten to 2500 with an
increase of ten objects between each measurement. At each stage was run 32 times
and the results were averaged to give smoother graphs. Two benchmarks was run,
one with an even distribution of objects and one with an uneven distribution.

@@

\subsubsection{Analysis}
There's virtually no difference in performance between the even and uneven
insertion times. This is because the codepath taken for each insertion is the
same, save for a different hash and bucket index. Because there is only a fixed
amount of boxes able to be stored in each grid cell in both grid types the
insertion time is run in $O(1)$, constant time. The case isn't the same for the
quadtree since its a recursive algorithm. Each node has four sub-nodes which get
traversed each insertion leading to a $O(\log_4 n)$ runtime since each step in
the algorithm eliminates three of the sub-nodes. Once the correct node has been
found the insertion time is a constant $O(1)$ since it uses the same technique
as the other two.

@@

@@

@@

\section{Sources}
Nystrom, Robert (2011) \textit{Game Programming Patterns}

>>>>

    {\Huge \textbf{Spatial Partitioning Strategies}}\\[1.5cm]

@@

\begin{abstract}
    Spatial partitioning is a widely used optimization strategy to efficiently
    optimize objects in space. It can be found in many applications, from
    simulating galaxies to video games.
    Checking collision between objects is an expensive operation that quickly
    becomes a bottleneck as the number of objects grow.
    This paper compares three different strategies, alongside a naive solution.
    The aim is to evaluate the use cases for the different strategies.
    Each strategy was implemented to fit the same interface to make it as fair
    of a comparison as possible.
    The results show how quadtrees out performs every other method when
    the objects are evenly and uneven distributed as the number of objects grow.
\end{abstract}

@@

\subsection{Vector}
A vector refers to a dynamic array, not the mathematical concept of a vector.
To allocate an array on the stack you need to know the size of it at compile
time which means vectors have to live on the heap. The vector works by having a
capacity which slowly fills up. Once it is full it resizes. This operation can
be quite expensive since it needs to request new memory from the operating
system and move all existing items into the new memory region.

@@

\section{Results \& Analysis}
This section presents the results and offers an analysis of the benchmarking
done on the three strategies - quadtrees, spatial hashing and static grids. The
metrics measured was average runtime: the sum of all operations needed to run
the program. Insertions: how long it took to insert all the objects into the
data structure. Collision testing: total time spent checking objects against
each other, which includes querying. Querying: average time spent retrieving
objects from the data structure. Clearing: how long it takes to reset the data
structure making it ready for rebuilding. Each metric was measured with a range
of object count, labeled as 'objects' in the graphs, from ten to 2500 with an
increase of ten objects between each measurement. At each stage was run 32 times
and the results were averaged to give smoother graphs. Two benchmarks was run,
one with an even distribution of objects and one with an uneven distribution.

@@

\subsubsection{Analysis}
There's virtually no difference in performance between the even and uneven
insertion times. This is because the codepath taken for each insertion is the
same, save for a different hash and bucket index. Because there is only a fixed
amount of objects able to be stored in each grid cell in both grid types the
insertion time is run in $O(1)$, constant time. The case isn't the same for the
quadtree since its a recursive algorithm. Each node has four sub-nodes which get
traversed each insertion leading to a $O(\log_4 n)$ runtime since each step in
the algorithm eliminates three of the sub-nodes. Once the correct node has been
found the insertion time is a constant $O(1)$ since it uses the same technique
as the other two.

@@

\subsubsection{Analysis}
In both cases the quadtree query has almost exactly the same time with the same
$O(\log_4 n)$ runtime complexity. The other two, however, differ quite a bit
time wise. This discrepancy is due to the use of vectors. Since a vector is used
to return the overlapping objects the act of building this vector becomes more
expensive as the amount of objects increases. Since there are more objects near
each other in the unevenly distributed objects case there will naturally be more
objects added to the vector. Quadtrees do not suffer from the same weakness
since it has a limit on the amount of objects stored in each leaf meaning all
queries has an upper limit to the number of objects, eliminating the need for
vector resizing.

The reason for the jaggedness in the graphs is because the execution time is so
low the benchmark picks up on other things like the OS scheduler and other
application doing work in the background taking away resources from the
program.

@@

\subsection{Analysis}
Just like with querying the same thing happens with collisions. Because there
are more objects close together there's a higher number of collision checks
taking place. Since the quadtree limit the amount of objects in a cell the
runtime doesn't change much between the two cases.

@@

\section{Sources}
Nystom, Robert (2014) \textit{Game Programming Patterns}, genever benning

Nystom, Robert (2021) \textit{Crafting Interpreters}, genever benning
